---
title: "Bayesian Week 6"
author: "Nikols"
date: "2026-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bayesian Statistics Practice Examples
# https://a-little-book-of-r-for-bayesian-statistics.readthedocs.io/en/latest/src/bayesianstats.html 
# First half of this document will be following along this website
```{r}
library("LearnBayes")
library(readr)
library(readxl)
```

```{r}
#Estimating Proportions: People who like chocolate
quantile1 <- list(p=0.5, x=0.85) #this specifies a median of the prior is 0.85
quantile2 <- list(p=0.99999, x=0.95) #the 99.999th percentile of the prior is 0.95
quantile3 <- list(p=0.00001, x=0.60) #the 0.001st percentile of the prior is 0.60
findBeta <- function(quantile1,quantile2,quantile3)
{
     # find the quantiles specified by quantile1 and quantile2 and quantile3
     quantile1_p <- quantile1[[1]]; quantile1_q <- quantile1[[2]]
     quantile2_p <- quantile2[[1]]; quantile2_q <- quantile2[[2]]
     quantile3_p <- quantile3[[1]]; quantile3_q <- quantile3[[2]]

     # find the beta prior using quantile1 and quantile2
     priorA <- beta.select(quantile1,quantile2)
     priorA_a <- priorA[1]; priorA_b <- priorA[2]

     # find the beta prior using quantile1 and quantile3
     priorB <- beta.select(quantile1,quantile3)
     priorB_a <- priorB[1]; priorB_b <- priorB[2]

     # find the best possible beta prior
     diff_a <- abs(priorA_a - priorB_a); diff_b <- abs(priorB_b - priorB_b)
     step_a <- diff_a / 100; step_b <- diff_b / 100
     if (priorA_a < priorB_a) { start_a <- priorA_a; end_a <- priorB_a }
     else                     { start_a <- priorB_a; end_a <- priorA_a }
     if (priorA_b < priorB_b) { start_b <- priorA_b; end_b <- priorB_b }
     else                     { start_b <- priorB_b; end_b <- priorA_b }
     steps_a <- seq(from=start_a, to=end_a, length.out=1000)
     steps_b <- seq(from=start_b, to=end_b, length.out=1000)
     max_error <- 10000000000000000000
     best_a <- 0; best_b <- 0
     for (a in steps_a)
     {
        for (b in steps_b)
        {
           # priorC is beta(a,b)
           # find the quantile1_q, quantile2_q, quantile3_q quantiles of priorC:
           priorC_q1 <- qbeta(c(quantile1_p), a, b)
           priorC_q2 <- qbeta(c(quantile2_p), a, b)
           priorC_q3 <- qbeta(c(quantile3_p), a, b)
           priorC_error <- abs(priorC_q1-quantile1_q) +
                           abs(priorC_q2-quantile2_q) +
                           abs(priorC_q3-quantile3_q)
           if (priorC_error < max_error)
           {
             max_error <- priorC_error; best_a <- a; best_b <- b
           }
       }
    }
    print(paste("The best beta prior has a=",best_a,"b=",best_b))
}

findBeta(quantile1,quantile2,quantile3)
```
# Beta prior
## This output means that the most appropriate prior to use for the proportion of people who like chocolate is a Beta prior with a=52.22, b=9.52 AKA a Beta(52.22, 9.52) prior. To plot this, use the curve() function: 
```{r}
curve(dbeta(x,52.22,9.52105105105105))
#Curve output represents initial beliefs that the peak of the distribution would be around 0.85, and that almost the entire density lies between 0.68-0.97. 
```
#Calculating the Likelihood Function for a Proportion
## Used to estimate a proportion with a small dataset, for example a survey of 50 people to see if they like chocolate -> data used to calculate the conditional p.m.f (probability mass function) of the proportion given the observed data = the likelihood function. 
#This is used to estimate how likely the possible values of the proportion are given the observed data.
```{r}
calcLikelihoodForProportion <- function(successes,total)
{
     curve(dbinom(successes,total,x)) # plots the likelihood
}
calcLikelihoodForProportion(45,50) #this hypothetical data would be a survey of 50 people -> 45 say they like chocolate (45 "successes") -> curve shows that the peak of the likelihood distribution is equal to 0.9, which is equivalent to 45/50 -> means that the most likely value of the proportion, given the observed data, is 0.9. 
```
#Calculating Posterior Distribution for a Proportion
# If you are trying to estimate a proportion and have a prior distribution and have collected some data, you can calculate the likelihood function for the proportion given the data.
# After observing the data you may wish to update the prior distribution for the proportion -> calculate the conditional distribution of the proportion given the data and the prior. This is called the posterior distribution for the proportion. 
# Summarizes what is known about the proportion after the data has been observed and combines the information from the prior and the data. 
# Chocolate example: Beta(52.22, 9.42) prior distribution; data from the survey found 45/50 people like chocolate. We can calculate the posterior distribution for the proportion given the prior and data using the function calcPosteriorForProportion() as follows:
```{r}
calcPosteriorForProportion <- function(successes, total, a, b)
  {
     # Adapted from triplot() in the LearnBayes package
     # Plot the prior, likelihood and posterior:
     likelihood_a = successes + 1; likelihood_b = total - successes + 1
     posterior_a = a + successes;  posterior_b = b + total - successes
     theta = seq(0.005, 0.995, length = 500)
     prior = dbeta(theta, a, b)
     likelihood = dbeta(theta, likelihood_a, likelihood_b)
     posterior  = dbeta(theta, posterior_a, posterior_b)
     m = max(c(prior, likelihood, posterior))
     plot(theta, posterior, type = "l", ylab = "Density", lty = 2, lwd = 3,
          main = paste("beta(", a, ",", b, ") prior, B(", total, ",", successes, ") data,",
          "beta(", posterior_a, ",", posterior_b, ") posterior"), ylim = c(0, m), col = "red")
     lines(theta, likelihood, lty = 1, lwd = 3, col = "blue")
     lines(theta, prior, lty = 3, lwd = 3, col = "green")
     legend(x=0.8,y=m, c("Prior", "Likelihood", "Posterior"), lty = c(3, 1, 2),
          lwd = c(3, 3, 3), col = c("green", "blue", "red"))
     # Print out summary statistics for the prior, likelihood and posterior:
     calcBetaMode <- function(aa, bb) { BetaMode <- (aa - 1)/(aa + bb - 2); return(BetaMode); }
     calcBetaMean <- function(aa, bb) { BetaMean <- (aa)/(aa + bb); return(BetaMean); }
     calcBetaSd   <- function(aa, bb) { BetaSd <- sqrt((aa * bb)/(((aa + bb)^2) * (aa + bb + 1))); return(BetaSd); }
     prior_mode      <- calcBetaMode(a, b)
     likelihood_mode <- calcBetaMode(likelihood_a, likelihood_b)
     posterior_mode  <- calcBetaMode(posterior_a, posterior_b)
     prior_mean      <- calcBetaMean(a, b)
     likelihood_mean <- calcBetaMean(likelihood_a, likelihood_b)
     posterior_mean  <- calcBetaMean(posterior_a, posterior_b)
     prior_sd        <- calcBetaSd(a, b)
     likelihood_sd   <- calcBetaSd(likelihood_a, likelihood_b)
     posterior_sd    <- calcBetaSd(posterior_a, posterior_b)
     print(paste("mode for prior=",prior_mode,", for likelihood=",likelihood_mode,", for posterior=",posterior_mode))
     print(paste("mean for prior=",prior_mean,", for likelihood=",likelihood_mean,", for posterior=",posterior_mean))
     print(paste("sd for prior=",prior_sd,", for likelihood=",likelihood_sd,", for posterior=",posterior_sd))
}
calcPosteriorForProportion(45, 50, 52.22, 9.52)
```
# Result: a peak in the posterior is seen on the graph approximately halfway between the peaks of the likelihood given the survey and the prior -> indicates that the data contribute roughly equally to the posterior. Note that the likelihood has been scaled so that the area underneath it is also 1, making it easy to compare the likelihood with the prior and the posterior.


#Dice Lab Example
```{r}
#plotting the overall probability of rolling each possible number on a d4, d6, d8, d10, d12, and d20
#Indicates that a d4 has a 25% chance of rolling any integer from 1-4, a d20 has a 5% chance of rolling any integer from 1-20, etc. 
par(mfrow=c(2,3), mar=c(4,4,1,0))
barplot(c(rep(1/4,4), rep(0,20-4)), ylim=c(0,1), names=c(1:20),  ylab="Probability", main="d4", xlab="dice roll")
barplot(c(rep(1/6, 6), rep(0, 20-6)), ylim=c(0, 1), names=c(1:20),
ylab="Probability", main="d6", xlab="dice roll")
barplot(c(rep(1/8, 8), rep(0, 20-8)), ylim=c(0, 1), names=c(1:20),
ylab="Probability", main="d8", xlab="dice roll")
barplot(c(rep(1/10, 10), rep(0, 20-10)), ylim=c(0, 1), names=c(1:20),
ylab="Probability", main="d10", xlab="dice roll")
barplot(c(rep(1/12, 12), rep(0, 20-12)), ylim=c(0, 1), names=c(1:20),
ylab="Probability", main="d12", xlab="dice roll")
barplot(c(rep(1/20, 20)), ylim=c(0, 1), names=c(1:20),
ylab="Probability", main="d20", xlab="dice roll")
```
```{r}
LynnRolls <- c(1, 2, 2, 3, 3, 4)
MattRolls <- c(1, 2, 3, 7, 7, 8)
```
```{r}
#Likelihood of Lynn rolling these numbers on a d4:
#L(d4 | LynnRolls) = P(LynnRolls | d4) because likelihood and probability are inversely related. The right hand side of the equation is the same as multiplying all of these rolls together because they are independent of one another, therefore:
#P(1, 2, 2, 3, 3, 4 | d4) = P(1 | d4)*P(2 | d4)*P(2 | d4)*P(3 | d4)*P(3 | d4)*P(4 | d4)
#Because each face comes up 25% of the time on a four sided die (see plot above), these probabilities are all equivalent to 0.25, therefore:
#P(1, 2, 2, 3, 3, 4 | d4) =  P(1 | d4)*P(2 | d4)*P(2 | d4)*P(3 | d4)*P(3 | d4)*P(4 | d4)
# = 0.25*0.25*0.25*0.25*0.25*0.25, or: 
(0.25)^6
# Output=0.00024
log((0.25)^6)
# Log Likelihood: -8.31
```
```{r}
#Likelihood of Lynn's rolls if using a d20: 
#The math is the same, except the probability of each integer being rolled on a d20 is 5% instead of 25%
##P(1, 2, 2, 3, 3, 4 | d20) = P(1 | d20)*P(2 | d20)*P(2 | d20)*P(3 | d20)*P(3 | d20)*P(4 | d20)
#P(1, 2, 2, 3, 3, 4 | d20)=
(0.05)^6
#Output: 1.5625e-08
log((0.05)^6)
#Log Likelihood: -17.97
#Compared to the d4 likelihood, the likelihood of rolling these numbers on a d20 is lower, so it is more likely that Lynn rolled her numbers on a d4 than a d20.
```
```{r}
#Likelihood of Matt's rolls on a d4 = 0, because he has rolls above 4, making the probability: 
# L(d4 | 1, 2, 3, 7, 7, 8) = P(1, 2, 3, 7, 7, 8 | d4)
# = P(1 | d4)*P(2 | d4)*P(3 | d4)*P(7 | d4)*P(7 | d4)*P(8 | d4)
# = 0.25 * 0.25 * 0.25 * 0 * 0 * 0
(0.25)^3 * (0)^3
# Output = 0
# Meanwhile for a d20: L(d20 | 1, 2, 3, 7, 7, 8) = P(1, 2, 3, 7, 7, 8 | d20)
(0.05)^6 
#Output: 1.5625e-08
#D20 is, by definition, more likely than a d4 for Matt's rolls. 
```
```{r}
#Plotting the likelihood of both sets of rolls on each die:
dice = c(4,6,8,10,12,20)
x <- dice
probdice = 1/x #probability of each die roll on each die
likelynn = probdice^6
likematt=likelynn
likematt[1:2]=0 
likematt
log10(likematt)
likelynn
log10(likelynn)
```
```{r}
plot(dice, log10(likelynn), col="red", type="b", xlab="sides", ylab="Log Likelihood")
points(dice, log10(likematt), col="blue", type="b", xlab="sides", ylab="Log Likelihood", pch=0, lty=2) 
legend("topright", c("Lynn", "Matt"), col=c("red", "blue"), pch=c(1, 0), lty=c(1, 2))
```
#Bayesian Inference
#Posterior probabilities add in priors/allow you to add in informatino you already know about a system in order to influence your final corrected likelihood. Posterior probabilities are approximately the prior probabilities multiplied by the likelihood of your data. 
#Lynn and Matt's rolls with priors (in this case, number of each die in their personal collections = 100):
#Lynn has 10, 60, 10, 10, 10, and 10 of d4s, d6s, d8s, d10s, d12s, and d20s, respectively.
#Matt has 10, 10, 10, 10, 30, and 30 of d4s, d6s, d8s, d10s, d12s, and d20s, respectively.
#Matt's priors:
```{r}
dice = c(4,6,8,10,12,20)
x <- dice
probdice = 1/x #probability of each die roll on each die
likematt = probdice^6 #Matt's likelihoods 
priormatt=c(0.1, 0.1, 0.1, 0.1, 0.3, 0.3)
posteriormatt=likematt*priormatt
flatprior = rep(1/6,6)
flatmatt = likematt*flatprior
```
#Lynn's priors:
```{r}
dice = c(4,6,8,10,12,20)
x <- dice
probdice = 1/x #probability of each die roll on each die
likelynn = probdice^6
priorlynn=c(0.1, 0.6, 0.1, 0.1, 0.1, 0.1)
posteriorlynn=likelynn*priorlynn
flatlynn=likelynn*flatprior
```
#Plotting Matt and Lynn's priors:
```{r}
par(mfrow=c(1,2))
plot(dice, log10(likematt), type="b", col="blue", pch=0, lty=2, ylim=c(-10, -4), xlim=c(8, 20))
points(dice, log10(flatmatt), type="b", col="blue", pch=1, lty=1, ylim=c(-10, -4), xlim=c(8, 20))
points(dice, log10(posteriormatt), type="b", col="blue", pch=2, lty=3, ylim=c(-10, -4), xlim=c(8, 20))
legend("bottomleft", c("likelihood", "flatprior", "posterior"), pch=c(0, 1, 2), lty=c(2, 1, 3), col="blue")

#have to change this to re-include d4 and d6
plot(dice, log10(likelynn), type="b", col="red", pch=0, lty=2, ylim=c(-10, -4), xlim=c(8, 20))
points(dice, log10(flatlynn), type="b", col="red", pch=1, lty=1, ylim=c(-10, -4), xlim=c(8, 20))
points(dice, log10(posteriorlynn), type="b", col="red", pch=2, lty=3, ylim=c(-10, -4), xlim=c(8, 20))
legend("bottomleft", c("likelihood", "flatprior", "posterior"), pch=c(0, 1, 2), lty=c(2, 1, 3), col="red")
```
# Second half: following along with the Dice lab sample data
# A 6 sided-die is called a d6, and people often collect polyhedral dice in standard sizes. Those sizes are d4, d6, d8, d10, d12, and d20, where dN is shorthand for a dice with N sides.
# You are following a table of six DND players where they are very bad at writing down which dice they roll, but they know exactly how many of each type of die they have. For each of the following sets of dice rolls, do a Bayesian analysis. Calculate and plot the log-likelihood and the posterior in R for all six dice types. Use the values in Table 2 as the basis for your priors.
```{r}
#number of rolls for each individual
BenjiRolls <- c(8, 8, 5, 16, 6, 8)
MarcusRolls <- c(12, 8, 2, 1, 4, 7)
CarrieRolls <- c(5, 8, 7, 3, 2, 6)
LucyRolls <- c(3, 5, 2, 3, 4, 3)
```
```{r}
NumberDice <- read_excel("NumberDiceTable2.xlsx")
View(NumberDice)
```
```{r}
#Plotting the likelihood of both sets of rolls on each die:
dice = c(4,6,8,10,12,20)
x <- dice
probdice = 1/x #probability of each die roll on each die
likeBenji = probdice^6
likeBenji[1:5]=0
likeMarcus = probdice^6
likeMarcus[1:4]=0
likeCarrie = probdice^6
likeCarrie[1:2]=0
likeLucy = probdice^6
likeLucy[1]=0

log10(likeBenji)
log10(likeMarcus)
log10(likeCarrie)
log10(likeLucy)
```
#Plotting each likelihood NOT taking into account how many dice each of them have
```{r}
par(mfrow=c(2,2), mar=c(4,4,1,0))
plot(dice, log10(likeBenji), type="b", xlab="sides", ylab="Log Likelihood", main="Benji's Rolls", ylim=c(-9, -4))
plot(dice, log10(likeMarcus), type="b", xlab="sides", ylab="Log Likelihood", main="Marcus's Rolls", ylim=c(-9, -4))
plot(dice, log10(likeCarrie), type="b", xlab="sides", ylab="Log Likelihood", main="Carrie's Rolls", ylim=c(-9, -4))
plot(dice, log10(likeLucy), type="b", xlab="sides", ylab="Log Likelihood", main="Lucy's Rolls", ylim=c(-9, -4))
```
#Now taking into account how many of each die they have (data is in the file NumberDice)

```{r}

```

