---
title: "Week6isolation"
author: "Nikols"
date: "2026-10-01"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Isolation data: Find the best fit model using response and interacting response variables.
##I'm still working on this but I wanted to put it out there that both of the following linear models are a good fit for the data: area + isolation and area * isolation as the predictor variables. I don't know if some parameters have to be adjusted because area is in km squared and isolation is in km 
```{r}
library(readr)
library(AICcmodavg)
library(car)
Isolation <- read_csv("Isolation.csv")
```
```{r}
#Linear models for variables where the response is incidence (0=bird does not breed there, 1=bird found). Explanatory variables are the area of the island (km2) and isolation of the island (distance from the mainland, km)
iso0 <- lm(incidence ~ 1, data=Isolation)
iso1 <- lm(incidence ~ area, data=Isolation)
iso2 <- lm(incidence ~ isolation, data=Isolation)
iso3 <- lm(incidence ~ area + isolation, data=Isolation)
iso4 <- lm(incidence ~ area * isolation, data=Isolation)
```

```{r}
#AICc Values
AICc(iso0)
AICc(iso1)
AICc(iso2)
AICc(iso3)
AICc(iso4)
```
```{r}
#Delta AICc values
(dAICC1 <- AICc(iso0) - AICc(iso4))
(dAICC2 <- AICc(iso1) - AICc(iso4))
(dAICC3 <- AICc(iso2) - AICc(iso4))
(dAICC4 <- AICc(iso3) - AICc(iso4)) 
(dAICC5 <- AICc(iso4) - AICc(iso4))
```

```{r}
par(mfrow = c(1, 2))
with(Isolation, (Boxplot(area, incidence)))
with(Isolation, (Boxplot(isolation, incidence)))
```
```{r}
summary(iso3)
summary(iso4)
```

