---
title: "Week6"
author: "Nikols"
date: "2026-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Cell Dataset
#Percent survival must be the only response variable
```{r}
## Perform model selection on the Cell_Inhibition_Schoenfeld.csv dataset.
# Investigate the relative importance of Inhibitor, Cell Line, VHL status, and Concentration of Inhibitor on Percent Survival. To do this properly, you will need to do the following:
# Use the lm() function with percent survival as the only response variable each time.
# Model just the intercept.
# Then, model at least 10 different combinations of those four predictors. At least one model must have an interaction.
# Test each model with aicc using the AICcmodavg package in R.
# After you are done with all models, calculate delta AICC for all of them.
# Describe your justification of why you chose those models. Did you start with the full model and work your way down? Did you start with univariate models and add predictors?
# Which model(s) explained the data the best?
# Take a summary of the best model(s) and describe which features are significant.
# What is your interpretation of the best model?
```
```{r}
library(readr)
cell <- read_csv("Cell_Inhibition_Schoenfeld.csv")
View(Cell_Inhibition_Schoenfeld)
library(AICcmodavg) 
```
```{r}
#Inhibitor (column=Inhibitor), Cell Line (col=`Cell Line`), VHL status (col=4), and Concentration of Inhibitor (col=`Concentration uM`) on Percent Survival(col=PercentSurvival)
cell1 <- lm(cell$PercentSurvival ~ 1)
cell2 <- lm(cell$PercentSurvival ~ cell$`Concentration uM` + cell$Inhibitor + cell$`Cell Line` + cell$VHL)
cell3 <- lm(cell$PercentSurvival ~ cell$`Concentration uM`*cell$Inhibitor + cell$`Cell Line` + cell$VHL)
cell4 <- lm(cell$PercentSurvival ~ cell$`Concentration uM`*cell$Inhibitor*cell$`Cell Line`)
cell5 <- lm(cell$PercentSurvival ~ cell$`Concentration uM`*cell$Inhibitor)
cell6 <- lm(cell$PercentSurvival ~ cell$`Concentration uM`)
cell7  <- lm(cell$PercentSurvival ~ cell$Inhibitor)
cell8 <- lm(cell$PercentSurvival ~ cell$`Cell Line`)
cell9 <- lm(cell$PercentSurvival ~ cell$VHL)
cell10 <- lm(cell$PercentSurvival ~ cell$`Cell Line` * cell$VHL)
```
```{r}
#AICc Values
AICc(cell1)
AICc(cell2)
AICc(cell3)
AICc(cell4)
AICc(cell5)
AICc(cell6)
AICc(cell7)
AICc(cell8)
AICc(cell9)
AICc(cell10)
```
#Answers:
#I chose these models because...
# Out of the ten models included, the best fit was the model lm(cell$PercentSurvival ~ cell$`Concentration uM` + cell$Inhibitor + cell$`Cell Line` + cell$VHL), the complete multivariate model with all four potential predictors included but wherein none of the models had an interaction term with one another. The next best model contained 
